---
title: "Class 08"
author: "Saira Cazarez PID: A69034931"
format: pdf
---

## Breast Cancer FNA data 
```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)

head(wisc.df)
```

> Q1. How many rows/patients/subjects. 

```{r}
nrow(wisc.df)
```
> Q2. How many M (cancer) B (healthy) patients 

```{r}
table(wisc.df$diagnosis)
```


Be sure to remove this diagnosis column from our data to analyze 

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- as.factor(wisc.df$diagnosis)
```
> Q3.How many variables/features in the data are suffixed with _mean?

```{r}
length( grep("_mean", colnames(wisc.data)))
```
```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

## Principal Component Analysis 

We want to scale our data before PCA by setting the `scale=True`argument! 

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
```


How much variance captured in each PC?
```{r}
summary(wisc.pr)
```
How much variance captured in each PC? 
```{r}
x <- summary(wisc.pr)
x$importance 
```
```{r}
plot(x$importance[2,], typ="b")
```
> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The plot is very difficult to understand because there are so many factors that are overlapping making the plot difficult to read. 

```{r}
biplot(wisc.pr)
```

```{r}
attributes(wisc.pr)
```

```{r}
head(wisc.pr$x)
```
My main PC results figure

```{r}

plot(wisc.pr$x , col = diagnosis , 
     xlab = "PC1", ylab = "PC2")

```

```{r}
plot(wisc.pr$x[,c(1,3) ], col =diagnosis, 
     xlab = "PC1", ylab = "PC3")
```

```{r}
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

library(ggplot2)

ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```
## Clustering 
Try to cluster the `wisc.data`. 
```{r}
km <- kmeans(wisc.data, centers= 2)
table(km$cluster)
```
```{r}
d <- dist (wisc.data)
hc <- hclust(d)
plot(hc)
```

```{r}
grps <- cutree(hc, k=3)
table(grps)
```

## Cluster in PC space 

In other words use my PCA results as a baisis of clustering. 

```{r}
d <- dist(wisc.pr$x[,1:3])
hc <- hclust(d, method="ward.D2")
plot(hc)
```


Cut this tree to yield 2 groups/clusters 
```{r}
grps <- cutree(hc, k=2)
table(grps)
```

Compare to my expert M and B `diagnosis`
```{r}
table(diagnosis)
```

```{r}
table(diagnosis, grps)
```


```{r}
pve <- (wisc.pr$sdev^2) / sum(wisc.pr$sdev^2)


plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```


```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```
> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```


```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
plot(wisc.hclust)
abline(h = 19, col = "red", lty = 2)

```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```
> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
wisc.hclust <- hclust(data.dist, method = "ward.D2")
plot(wisc.hclust)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```


```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)
```

```{r}
g <- relevel(g,2)
levels(g)
```
```{r}
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

Patient 2 should be prioritized because they have more malignant type cells. 
