---
title: "Class 15: Mini Project: Investigating Pertussis Resurgence"
author: "Saira Cazarez PID: A69034931"
format: gfm
---


Pertussis, a.k.a Whooping Cough, is a highly contagous lung infection caused by the *B. Pertussis*


The CDC tracks Pertussis case numbers and can be accessed[here](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

We need to "script" this data so we do stuff with it in R. Let's try the **datapasta** package to do this. 


```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L, 2024L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044, 23544)
       )
```


Let's plot year vs cases to see the trend over time in the US. 

```{r}
library(ggplot2)

baseplot <- ggplot(cdc)+ 
  aes(year, cases)+
  geom_point()+
  geom_line()
```


Let's add the date of wP vaccine roll out completion (1946) and the switch to the new aP vaccine (1996). 
```{r}
baseplot + 
  theme_bw()+
  geom_vline(xintercept= 1946, col="blue")+
  geom_vline(xintercept= 1996, col="red")+
  geom_vline(xintercept= 2020, col="gray")+
  geom_vline(xintercept= 2004)

```

##CMI-PB (Computational Mondels of Immunity- Pertussis Boost)

This project collects and makes freely available data about the immune response to Pertussis vaccination. 

You can access the data via API which resturns JSON format (key:vale pairs). 

We can use the **jsonlite** package and it's `read_json()` function. 

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5/subject",
                      simplifyVector = TRUE)
```


Let's have a wee peak and explore this 
```{r}
head(subject)
```

> Q. How many subjects do we have ? 

```{r}
nrow(subject)
```

>Q. How many make/female do we have? 

```{r}
table(subject$biological)
```

> Q. How many wP and aP do we have? 

```{r}
table(subject$infancy_vac)
```


>Q. break down of biological sex and race? 

```{r}
table(subject$race, subject$biological_sex)
```

>Q. Does this break down reflect the US population? 

No 

```{r}
table(subject$dataset)
```



```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5/specimen" ,
                      simplifyVector = TRUE)

ab_titer <- read_json("https://www.cmi-pb.org/api/v5/plasma_ab_titer" ,
                      simplifyVector = TRUE)
```


```{r}
head(ab_titer)
```

```{r}
head(specimen)
```


We want merge or "join" these tables so we can have all the infor we need about a given antibody measurement. 
```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
```
```{r}
head(meta)
```

and one last join of `ab_titer` and `meta`

```{r}
abdata <- inner_join(ab_titer, meta)
```
```{r}
head(abdata)
```

```{r}
nrow(abdata)
```

```{r}
head(abdata)
```

```{r}
table(abdata$isotype)
```

```{r}
table(abdata$antigen)
```


Let's begin with IgG 
```{r}
igg <- filter(abdata, isotype=="IgG")
head(igg)
```
Make a boxplot of IgG antigen levels - this will be a plot of MFI vs antigen 
```{r}
library(ggplot2)

ggplot(igg)+
aes(MFI_normalised, antigen)+
  geom_boxplot()
```


```{r}
ggplot(igg)+
aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()
```

Ideally I would like to see how these Ab levels change iver time relative to the booster shot. 

```{r}
table(abdata$visit)
```
```{r}
ggplot(igg)+
aes(MFI_normalised, antigen)+
  geom_boxplot()+
  facet_wrap(~visit)
```



```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost, y=MFI_normalised, col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```


```{r}
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost, y=MFI_normalised, col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```








